Classless Java

### 2016-03-23 OOPSLA16/sections/Implementation.tex
Ref to the blog. Didn't find the author name.

%\footnote{ Rewriting libraries and
%  application in this language/model would result in more reusable
%  libraries, but it would be a taunting task.}
Creating a new language or language extension would be an
elegant way to illustrate the point of IB. However,
significant amounts of engineering would be needed to build a practical
language, and to achieve a similar level of integration and tool support
as Java. To have an approach that can both illustrate IB
programming and be practical, we have instead implemented Classless Java.
This section presents the Classless Java design.
Disciplined use of Classless Java (that is avoiding class
definitions as done in Section~\ref{}), illustrates what \emph{pure} IB is like.
However in Classless Java CB and IB programming can be mixed together, and all
the practical convenience of using existing Java libraries, the full
Java language, and IDE's are available in Classless Java.
The key to our implementation is the use of compilation agents, which
are supported in Java, and allow us to rewrite the Java AST just
before compilation. We discuss advantages and limitations of our approach.

\subsection{Compilation Agents}
Java supports compilation agents\cite{compilationagents}.
That is, Java libraries can interact with the Java compilation process,
acting as a man in the middle between the
generation of the abstract syntax tree and the Java Bytecode.
\marco{is this correct?}\bruno{my impression is that Lombok happens
  \emph{before} the AST is compiled (not between AST and bytecode generation)}
This process is facilitated by frameworks like Lombok~\cite{lombok}:
a Java library that aims at removing (or
reducing) Java boilerplate code via
annotations. Classless Java was created using Lombok.
Figure~\ref{fig:lombok} illustrates the
flow of \mixin annotation processing in Lombok.\bruno{Haoyuan, please
  explain the diagram.}
\haoyuan{Source code in a Java program is first parsed into its syntactic structure,
which we call abstract syntax tree (AST). The AST is passed to Lombok, and a corresponding
AST defined in Lombok is generated for the annotated node. An annotation like \mixin is declared
inside Lombok, together with an implementation of its (Eclipse or Javac) handler. The handler is
free to modify the information of the annotated node, or even inject new nodes (like methods, inner classes,
etc). Finally, Java compiler works on the modified AST to generate bytecode. During the compilation,
no source code is changed, and no new files are created, bytecode takes the only difference.}

@@

\paragraph{Advantages of Lombok}
The Lombok compilation agent has advantages with respect to alternatives like
pre-processors, or other Java annotation processors. In some sense
Lombok can be seen as a poor man's macro system~\cite{} for Java.
\bruno{Marco, do you agree with calling Lombok a poor man's macro system?}
\marco{No, it is as powerful as lisp macros, and have a bigger (may be even  a better) design.}
%this seems like a commercial spot
%\item Lombok is byte-code based instead of source-code based, which makes client
%  code concise and easy to maintain. Such code generation is performed at
%  compile time to modify bytecode.

@@

\paragraph{Tool support:}
Features written in Lombok integrate and are supported directly in the
language, and are often also supported by (most) tools.  For example in Eclipse, the processing is
performed transparently and the information of the interface from
compilation is captured in the ``Outline'' window.\bruno{Haoyuan: show
and explain figure here.} This includes all
the methods inside the interface as well as the generated ones. \haoyuan{In Figure~\ref{fig:screenshot},
\mixin generates \Q@of@ method in \Q@Point2D@, and \Q@of@, \Q@withX@, \Q@withY@ methods in \Q@Point3D@. These
methods are also visible to users, showing their types and modifiers in Outline window. Moreover, as a useful IDE feature,
the auto-completion also works for these newly generated methods.}

@@

\subsection{AST Reinterpretation in Classless Java}

Ofcourse, careless reinterpretation of the AST could still be
surprising for badly designed rewritings.  Classless Java reinterprets
the syntax with the sole goal of \emph{enhancing and completing code}:
we satisfy the behaviour of abstract methods; we add method
implementations; and we refine return types.  We consider this to be
quite easy to follow and reason about, since it is similar to what
happens in normal inheritance.  Refactoring operations like renaming
and moving should work transparently in conjunction with our
annotaiton, since they rely on the overall type structure of the
class, that we do not arbitrarily modify but just complete.

@@

\paragraph{Heir coherence:} Another form of guarantee that could be
useful in AST rewriting would be heir coherence. That is, interfaces
(and in general classes) inheriting the instrumented code are
well-typed if they were well-typed without the instrumentation.
Classless Java \emph{does not} guarantee heir coherence.  The reason
is that this would forbid adding any (default or abstract) method to
the annotated interfaces, including type refinement. Indeed consider
the following:

@@

Our automatic type refinement is a useful and convenient feature, but
not transparent to the heirs of the annotated interface.  They need to
be aware of the annotation semantics and provide the right type while
refining methods. To support heir coherence, Classless Java would need
to give up automatic type refinement, which is an essential part of IB programming.
However, the reader should note that Java libraries almost always break heir
coherence during evolution and still claim backward compatibility (false in
theory but statistically true in practice). In practice to break heir
coherence it is just enough to add any method to any
non final class of a Java library.  We think refining return
types breaks heir coherence ``less" than normal library evolution.

@@

\subsection{Limitations}\marco{ to reorganize with updated limitations}
\haoyuan{Should we only present the separate compilation and simple generics features in this (limitation) section?}
Our prototype implementation using Lombok has certain limitations:
\begin{itemize}
\item \haoyuan{The annotation \mixin does not support separate compilation yet. Using \mixin, all
  related interfaces have to appear in a single Java file. Therefore, changes to
  a single interface would require re-compiling the whole file. Nevertheless, to support separate compilation we
  studied an existing yet experimental feature \lstinline{@Delegate}, reused the name and modified its behavior.
  Now files in the same package can be compiled and \lstinline{@Delegate} has the same behavior as \mixin. A limitation is
  that users need to put references to the super types of the annotated type together with the annotation, for instance,
  \lstinline{@Delegate(types = Point2D.class)}. Another limitation is the methods generated by \lstinline{@Delegate} will
  not be visible in the Outline window, but they can still be auto-completed in Eclipse.}
\item At this stage our implementation only realizes the Eclipse handler and our
  experiments are all conducted in Eclipse. The implementation for
  \texttt{javac} is missing.
\item \haoyuan{The current implementation only supports simple generics, which means type parameters can be used, however, the generic
method typing is neither formalized in this paper nor implemented, but delegated to the Java compiler.}
\end{itemize}

>>>>

%\footnote{ Rewriting libraries and
%  application in this language/model would result in more reusable
%  libraries, but it would be a taunting task.}
Creating a new language or language extension would be an
elegant way to illustrate the point of IB. However,
significant amounts of engineering would be needed to build a practical
language, and to achieve a similar level of integration and tool support
as Java. To have an approach that can both illustrate IB
programming and be practical, we have instead implemented 
\mixin as an annotation over Java8, and a \emph{compilation agent}.
That is, the Classless Java style of programming
is supported by a Java8 library.

Disciplined use of Classless Java (that is avoiding class
definitions as done in Section~\ref{}), illustrates what \emph{pure} IB is like.
However, using \mixin, CB and IB programming can be mixed together, 
harvesting the practical convenience of using existing Java libraries, the full
Java language, and IDE's support.
The key to our implementation is the use of compilation agents, which
are supported in Java, and allow us to rewrite the Java AST just
before compilation. We discuss advantages and limitations of our approach.

\subsection{Compilation Agents}
Java supports compilation agents\cite{compilationagents}.
That is, Java libraries can interact with the Java compilation process,
acting as a man in the middle between the
generation of the abstract syntax tree and the Java Bytecode.

This process is facilitated by frameworks like Lombok~\cite{lombok}:
a Java library that aims at removing (or
reducing) Java boilerplate code via
annotations. \mixin was created using Lombok.
As you can see explained in Figure~\ref{fig:lombok}~\cite{neildo2011blog},
first java source code is parsed into an abstract syntax tree (AST).
The AST is then captured by Lombok:
each annotated node is passed to
the corresponding (Eclipse or Javac) handler. The handler is
free to modify the information of the annotated node, or even inject new nodes (like methods, inner classes,
etc). Finally, the Java compiler works on the modified AST to generate bytecode.
%Note how during the compilation,
%no source code is changed, and no new source/temporary files are created.

@@

\paragraph{Advantages of Lombok}
The Lombok compilation agent has advantages with respect to alternatives like
pre-processors, or other Java annotation processors.
Lombok offers in Java an expressive power similar to that of Scala/Lisp macros;
except of course, for the syntactic convenience of quote/unquote templating.

@@

\paragraph{Tool support:}
Features written in Lombok integrate and are supported directly in the
language, and are often also supported by (most) tools.  For example in Eclipse, the processing is
performed transparently and the information of the interface from
compilation is captured in the ``Outline'' window.
This includes all
the methods inside the interface as well as the generated ones.
In Figure~\ref{fig:screenshot},
\mixin generates \Q@of@ method in \Q@Point2D@, and \Q@of@, \Q@withX@, \Q@withY@ methods in \Q@Point3D@.
These methods are also visible to users, showing their types and modifiers in Outline window.
Moreover, as a useful IDE feature, the auto-completion also works for these newly generated methods.

@@

\subsection{\mixin AST Reinterpretation}

Of course, careless reinterpretation of the AST could still be
surprising for badly designed rewritings.  \mixin reinterprets
the syntax with the sole goal of \emph{enhancing and completing code}:
we satisfy the behaviour of abstract methods; we add method
implementations; and we refine return types.  We consider this to be
quite easy to follow and reason about, since it is similar to what
happens in normal inheritance.  Refactoring operations like renaming
and moving should work transparently in conjunction with our
annotation, since they rely on the overall type structure of the
class, that we do not arbitrarily modify but just complete.

@@

\paragraph{Heir coherence:} Another form of guarantee that could be
useful in AST rewriting would be heir coherence. That is, interfaces
(and in general classes) inheriting the instrumented code are
well-typed if they were well-typed without the instrumentation.
Our rewriting \emph{does not} guarantee heir coherence.  The reason
is that this would forbid adding any (default or abstract) method to
the annotated interfaces, including type refinement. Indeed consider
the following:

@@

Our automatic type refinement is a useful and convenient feature, but
not transparent to the heirs of the annotated interface.  They need to
be aware of the annotation semantics and provide the right type while
refining methods. To support heir coherence, we would need
to give up automatic type refinement, which is an essential part of IB programming.
However, the reader should note that Java libraries almost always break heir
coherence during evolution and still claim backward compatibility (false in
theory but statistically true in practice). In practice to break heir
coherence it is just enough to add any method to any
non final class of a Java library.  We think refining return
types breaks heir coherence ``less" than normal library evolution.

@@

\subsection{Limitations}
Our prototype implementation using Lombok has certain limitations:
\begin{itemize}
\item Eclipse has its own compiler (ejc), alternative to javac.
  Lombok allows to write handlers for either javac of ejc.
  At this stage our implementation only realizes the Eclipse handler and our
  experiments are all conducted in Eclipse. The implementation for
  the \texttt{javac} handler is still missing.
\item The current implementation supports generics in a simplified manner: 
type parameters can be used but the generic
method typing is neither formalized in this paper nor explicitly checked by our 
implementation, but simply  delegated to the Java compiler.

\item 
Lombok offers only limited/experimental support for separate compilation, that is
to access information of code defined in different files.
In the same spirit, we have a mature \mixin annotation, that
does not support separate compilation yet: it requires that all
  related interfaces have to appear in a single Java file.
We also offer a less polished annotation, 
that trying to reuse the (yet experimental) Lombock annotation \lstinline{@Delegate}, reused the name and modified its behavior.
  Now files in the same package can be compiled and \lstinline{@Delegate} has the same behavior as \mixin. A limitation is
  that users need to put references to the super types of the annotated type together with the annotation, for instance,
  \lstinline{@Delegate(types = Point2D.class)}. Another limitation is the methods generated by \lstinline{@Delegate} will
  not be visible in the Outline window, but they can still be auto-completed in Eclipse.
\end{itemize}
